---
title: D-statistics
description: >-
  Infer introgression and hybridization with D-statistics
author:
  - Per Unneberg
format:
  nbis-course-html:
    number-sections: false
exercise:
  dir: population-structure
  venv: e-population-structure
  data: monkeyflower/selection/large/vcftools-filter-bqsr/
  wget_opts: \-\-accept-regex all.variantsites* \-\-cut-dirs=7
---

<!-- markdownlint-disable MD041 -->

{{< include /exercises/_knitr.qmd >}}

{{< include /exercises/_rbuild.qmd >}}

{{< include /exercises/_rlibs.qmd >}}

```{r }
#| label: r-init-vars
#| echo: false
#| eval: true
#| cache: false
obj <- exercise_tools$population_structure$dstats
```

<!-- markdownlint-enable MD041 -->

## About

Populations that diverge often exchange migrants during the splitting
process. Populations may also come into secondary contact and
hybridize after being separated over long times. In this exercise, we
look at f- and D-statistics to detect patterns of hybridization and
migration (introgression), as implemented in the package {{< var
tools.admixtools2.fmt >}}.

{{< include /exercises/_subset.qmd >}}

:::{.callout-tip collapse=true}

## Intended learning outcomes

- use D-statistics to detect hybridization and introgression

:::

{{< include /exercises/_tools.qmd >}}

{{< include /exercises/_datasetup.qmd >}}

## On D-stastitics and f-statistics

D and f-statistics are tools used in explorations of
multiple-population histories, particularly in instances involving
migration or hybridization that form reticulate networks, as opposed
to bifurcating trees. D-statistics, also known as ABBA-BABA tests, are
designed to infer signs of introgression.

On the other hand, f-statistics are crucial for detecting admixture
from data. Specifically, f2 measures the total branch length between
populations, while f3 tests whether one population (P1) is an
admixture of two others (P2 and P3). Furthermore, f4 tests for the
independence between two pairs of populations (P1, P2 and P3, P4).

These statistics can be effectively implemented using the
comprehensive admixtools and admixtools2 packages. The packages offer
an array of capabilities, the extent of which will only be slightly
explored in this context. A more exhaustive overview can be found in
the [the admixtools
tutorial](https://uqrmaie1.github.io/admixtools/articles/admixtools.html).

## Data setup

First initialize some variables.

```{bash }
#| label: setup-datavars
#| echo: true
#| eval: true
VCF="variants.vcf.gz"
DATAPFX="monkeyflower_dstats"
```

```{r }
#| label: r-set-envvars
#| echo: false
#| eval: true
#| cache: false
Sys.setenv(
    "VCF" = "variants.vcf.gz",
    "DATAPFX" = "monkeyflower_dstats"
)
```

Here we need to invoke {{< var tools.r.fmt >}} from the get-go. Load
relevant libraries and register the variables.

```{r }
#| label: r-load-libraries
#| echo: true
#| eval: true
{{< include /exercises/_rlibs.R >}}
library(dplyr)
library(tibble)
library(admixtools)

VCF <- Sys.getenv("VCF")
DATAPFX <- Sys.getenv("DATAPFX")
```

{{< var tools.admixtools2.fmt >}} needs a population input file consisting of
two columns and is picky with regards to the format. Therefore, we
need to refactor the sampleinfo file and select relevant columns.

```{r }
#| label: make-population-file
#| echo: true
#| eval: true
sampleinfo <- read.csv("sampleinfo.csv") %>% rename(sample=SampleAlias) %>%
  mutate(species = as.factor(gsub("(ssp. |M. )", "", Taxon))) %>%
  mutate(species = as.factor(gsub(", ", "_", species))) %>%
  select(sample, species) %>%
  as_tibble
write.table(sampleinfo, file=paste0(DATAPFX, ".pop"),
            row.names=FALSE, col.names=FALSE,
            sep="\t", quote=FALSE)
```

## Preparing {{< var tools.admixtools2.fmt >}} input files

We need to convert our input VCF to a {{< var tools.plink.fmt >}} bed
file. Some of the steps are a bit fiddly and it can be tricky to infer
what the file formats look like from the documentation alone. Lucky
for you, we have prepared a template sequence of commands so you don't
have to look up the details. Fasten your seatbelt and read on!

### Convert VCF to plink ped format

To start with, {{< var tools.admixtools2.fmt >}} requires
integer-based chromosome names. We therefore reannotate the input VCF
with {{< var tools.bcftools.fmt >}}. Here, `chrmap.txt` is a tabular
file that maps old to new chromosome names.

```{bash }
#| label: bcftools-rename-chrom
#| echo: true
#| eval: true
for i in {3..4}; do echo -e "LG${i} ${i}"; done > chrmap.txt
bcftools annotate --rename-chrs chrmap.txt $VCF -o ${DATAPFX}.vcf.gz -O z -W tbi
```

We then convert the updated VCF to bed format.

```{bash }
#| label: plink2-vcf-to-ind
#| echo: true
#| eval: true
plink2 --vcf ${DATAPFX}.vcf.gz --allow-extra-chr \
       --set-missing-var-ids @:# \
       --make-bed \
       --out ${DATAPFX} > /dev/null 2>&1
```

Next, we make use of the `.pop` file we generated at the beginning to
add population information to the first column of the `.fam` file, as
well as modify the last column (phenotype) from `-9` (missing data) to
`1` (control); the latter needs to be set, else the sample is ignored.

```{bash }
#| label: csvtk-update-fam
#| echo: true
#| eval: true
mv ${DATAPFX}.fam ${DATAPFX}.orig.fam
csvtk join -H -t ${DATAPFX}.orig.fam ${DATAPFX}.pop -f "2;1" |
    csvtk cut -t -f 7,2,3,4,5,6 |
    csvtk replace -H -t -f 6 -r 1 > ${DATAPFX}.fam
head -n 3 ${DATAPFX}.fam
```

Next, we convert bed to ped format, which is the required input for
the EIGENSOFT package
[@patterson_PopulationStructureEigenanalysis_2006] in the next
section.

```{bash }
#| label: plink2-recode-bed-to-ped
#| echo: true
#| eval: true
plink2 --bfile ${DATAPFX} --recode 12 ped \
       --out ${DATAPFX} > /dev/null 2>&1
```

Here, `--bfile` specifies a set of files with suffixes `.bed`, `.bim`
and `.fam`, whereas `--recode` recodes nucleotides to numbers. Now you
should have a ped file, which is the required format for the next
step.

### Convert PED to EIGENSTRAT

The EIGENSOFT package
[@patterson_PopulationStructureEigenanalysis_2006] is a collection of
tools for performing eigenanalysis of population data. The `convertf`
program converts between different file formats. We here want to
convert from ped to EIGENSTRAT and thefore need to write a control
parameter file:

```{bash }
#| label: make-parameter-file
#| echo: true
#| eval: true
cat <<EOF  > par.PED.EIGENSTRAT
genotypename:    monkeyflower_dstats.ped
snpname:         monkeyflower_dstats.bim
indivname:       monkeyflower_dstats.fam
outputformat:    EIGENSTRAT
genotypeoutname: monkeyflower_dstats.geno
snpoutname:      monkeyflower_dstats.snp
indivoutname:    monkeyflower_dstats.orig.ind
EOF
```

The top three lines correspond to input files, whereas the bottom
three are output files. Note that we add an `.orig` tag to the
`indivoutname` as we (again) want to slightly modify the output later
on. With this parameter file, convertf is run as

```{bash }
#| label: convert-ped-to-eigenstrat
#| echo: true
#| eval: true
convertf -p par.PED.EIGENSTRAT > ${DATAPFX}.convertf.log 2>&1
```

We reformat the individual file to produce a desired output consisting
of sample, gender (unknown) and population group label ^[For more
information about the EIGENSOFT file formats, see
<https://github.com/DReichLab/EIG/tree/master/CONVERTF>].

```{bash }
#| label: massage-ind-file
#| echo: true
#| eval: true
csvtk space2tab ${DATAPFX}.orig.ind |
    csvtk cut -t -H -f 1,2,1 |
    csvtk replace -t -H -f 1 -p "^.*:" |
    csvtk replace -t -H -f 3 -p ":.*" > ${DATAPFX}.ind
head -n 3 ${DATAPFX}.ind
```

## Analyse data with {{< var tools.admixtools2.fmt >}}

Finally, after all the data juggling, we can start with the analyses!

### Prepare f2_blocks

The speedup of {{< var tools.admixtools2.fmt >}} compared with its
predecessor owes much to the efficient use made of $f_2$-statistics.
So much so in fact that it can be worthwhile to pre-calculate and
store $f_2$-statistics, as we do below.

```{r }
#| label: admixtools2-save-f2-blocks
#| echo: true
#| eval: true
f2_dir <- "f2"
extract_f2(DATAPFX, f2_dir, verbose=FALSE, overwrite=TRUE)
f2_blocks <- f2_from_precomp(f2_dir, verbose=FALSE)
```

`extract_f2` will compute and store blocked f2 statistics, where the
size of each block is 0.05cM by default.

### Plot average $f_2$-statistics

$f_2$-statistics measure the amount of genetic drift between two
populations (i.e., the total branch length). We can quickly summarize
the average $f_2$ over all blocks and make a plot:

```{r }
#| label: plot-matrix
#| echo: true
#| eval: true
apply(f2_blocks, 1:2, mean) %>% as_tibble %>%
  mutate(pop1=colnames(.), .before=1) %>%
  pivot_longer(!pop1, names_to="pop2") %>%
  ggplot(aes(x=pop1, y=pop2, fill=value)) + geom_tile() +
  scale_fill_viridis()
```

The structure of the plot could be improved by clustering on rows and
columns, but clearly population comparisons including *clevelandii*
have the highest values, as would be expected.

### Calculate {{< var fst >}}

$f_2$-statistics is mathematically related to {{< var fst >}}, and we
can easily obtain {{< var fst >}} estimates with the `fst` function
and plot the outcome, obtaining a plot similar to the one above:

```{r }
#| label: fst-from-f2
#| echo: true
#| eval: true
fst(f2_blocks) %>%
  ggplot(aes(x=pop1, y=pop2, fill=est)) + geom_tile() +
  scale_fill_viridis()
```

### f3 statistics

FIXME!

### f4 statistics (qpdstat)

The $f_4$ statistic compares four populations P1, P2, P3, P4 and tests
for independence (no admixture) between P1, P2 and P3, P4. Here we
generate all 56 possible pairs of 4-taxon tests, setting *clevelandii*
as the outgroup (P4).

```{r }
#| label: fig-f4-statistics
#| echo: true
#| eval: true
#| fig-width: 14
#| fig-height: 8
#| fig-cap:
#|   Genome-wide estimates of Patterson’s f4-statistic (this is similar
#|   to the D-statistic or ABBA-BABA test) for all possible pairs of
#|   4-taxon tests. The statistics have been sorted in ascending
#|   order.
allpops <- c("clevelandii", "grandiflorus", "aridus", "parviflorus",
             "longiflorus", "calycinus", "aurantiacus", "puniceus_yellow",
             "puniceus_red")
pops <- matrix(allpops[2:9][combn(8, 3)], , ncol=3, byrow=TRUE)[, 3:1]
pops <- cbind(pops, "clevelandii")
# f4 is synonym to qpdstat
res <- qpdstat(f2_blocks, pops)
res %>%  mutate(taxa=paste(pop1, pop2, pop3, sep=",")) %>%
  as_tibble %>% arrange(est) %>%
    mutate(taxa=factor(taxa, levels=as.character(taxa))) %>%
    ggplot(aes(x=taxa, y=est)) + geom_point() + xlab("Taxa (P1, P2, P3)") +
    ylab("D-statistic (P1, P2; P3, O)")
```

If no permutations have $f_4=0$, then there is no unadmixed tree that
can explain the data.

### qpgraph

FIXME!

```{r }
#| label: define-graph-wo-admixture
#| echo: true
#| eval: true
g <- matrix(c(
    'R', 'R', 'n1', 'n1', 'n2', 'n3', 'n3', 'n2', 'n4',
    'n4', 'n5', 'n6', 'n6', 'n5', 'n7', 'n7',
    'clevelandii', 'n1', 'grandiflorus', 'n2', 'n3',
    'parviflorus', 'aridus', 'n4', 'aurantiacus',
    'n5', 'n6', 'calycinus', 'longiflorus', 'n7',
    'puniceus-yellow', 'puniceus-red'), , 2) %>%  edges_to_igraph()
plot_graph(g)
```

```{r }
#| label: define-graph-admixture-long-y
#| echo: true
#| eval: true
g <- matrix(c(
    'R', 'R', 'n1', 'n1', 'n2', 'n3', 'n3', 'n2', 'n4',
    'n4', 'n5', 'n6', 'n6', 'n5', 'n7', 'n7', 'n6',
    'clevelandii', 'n1', 'grandiflorus', 'n2', 'n3',
    'parviflorus', 'aridus', 'n4', 'aurantiacus',
    'n5', 'n6', 'calycinus', 'longiflorus', 'n7',
    'puniceus-yellow', 'puniceus-red', 'n7'), , 2) %>%  edges_to_igraph()
plot_graph(g)
```

## Conclusions

That was a long exercise, but now you should have a feeling for how
you can apply f-statistics to your data. Well done!
