<!-- markdownlint-disable MD041 -->

::: {.callout-note collapse=true}

## Tools

:::{.panel-tabset}

#### Listing

```{r }
#| label: tool-listing
#| echo: false
#| eval: true
#| cache: false
#| output: asis
cat(format_listing(obj))
```

#### {{< fa server >}} PDC

Choose **one** of **Modules** and **Virtual environment** to access
relevant tools.

##### Modules

Execute the following command to load modules:

```{r }
#| label: r-pdc-load-modules
#| echo: false
#| eval: true
#| cache: false
#| output: asis
cat(format_modules(obj, bioinfo_tools=bioinfo_tools))
```

##### Virtual environment

Run the `pgip` initialization script and activate the `pgip`
default environment:

```bash
source {{< var pg_pgip_dir >}}/init.sh
pgip_activate
```

Then activate the `{{< meta exercise.venv >}}` environment:

```bash
# pgip_shell calls pixi shell -e {{< meta exercise.venv >}} --as-is
pgip_shell {{< meta exercise.venv >}}
```

#### {{< fa laptop >}} pixi

Copy the contents to a file `pixi.toml` in directory `{{< meta
exercise.dir >}}`, `cd` to directory and activate
environment with `pixi shell`:

```{r }
#| label: r-pixi-manifest
#| echo: false
#| eval: true
#| cache: false
#| output: asis
cat(format_pixi(obj, "{{< meta exercise.dir >}}"))
```

:::

:::

<!-- markdownlint-enable MD041 -->
